---
title: "Sanitation facilities (% of population with access)"
author: "Shuting Sun"
date: "10/5/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r pressure, echo=FALSE, warning=FALSE}

library(ggplot2)
library(WDI)
library(wbstats)
library(dygraphs)
library(dplyr)

rm(list=ls()) # clear environment

indicators_sanitation = 'SH.STA.SMSS.ZS'
indicators_population = 'SP.POP.TOTL'

countries_incomelevel <- function(inc) {  # a function, given an income level, get the 'iso3c' list of all the countries in that income level
  df_countries = wbcountries()
  countries = df_countries[which(df_countries$incomeID == inc), "iso3c"]
  return(countries)
}

dataset_incomelevel <- function(inc){  # a function, given an income level, return the data set with the safely managed sanitation facilities (%) each year
data_LIC = WDI(indicator=c(indicators_sanitation, indicators_population), 
               country = countries_incomelevel(inc),  start=1960, end=2018) # get the sanitaion and population data (1960 to 2018)

data_clean = data_LIC[complete.cases(data_LIC), ] # only keep the rows with both population data and sanitation data

data_clean$total <- data_clean$SH.STA.SMSS.ZS * data_clean$SP.POP.TOTL # total = population * sanitation(%)

total_population = aggregate(SP.POP.TOTL ~ year, data_clean, sum) # group the dataframe by "year", and sum the "population"
total_sanitation = aggregate(total ~ year, data_clean, sum) # group the dataframe by "year", and sum the "total"

data_sum = merge(x = total_population, y = total_sanitation, by = "year", all = TRUE) # merge the above two dataset

data_sum$trend <- data_sum$total / data_sum$SP.POP.TOTL # here is the weighted arithmetic mean of the sanitation (%population) data
data_sum$IncomeLevel <- rep(inc,nrow(data_sum)) # add an IncomeLevel column
return(data_sum)
}

data1 = dataset_incomelevel("LIC")
data2 = dataset_incomelevel("LMC")
data3 = dataset_incomelevel("UMC")
data4 = dataset_incomelevel("HIC")

data1_select = select(data1, year, trend)
names(data1_select)[names(data1_select) == 'trend'] <- 'LIC'
data2_select = select(data2, trend)
names(data2_select)[names(data2_select) == 'trend'] <- 'LMC'
data3_select = select(data3, trend)
names(data3_select)[names(data3_select) == 'trend'] <- 'UMC'
data4_select = select(data4, trend)
names(data4_select)[names(data4_select) == 'trend'] <- 'HIC'
data_select =cbind(data1_select, data2_select, data3_select, data4_select)
# xts(data_select$trend, as.Date(data_select$year, format='%m/%d/%Y')
dygraph(data_select, main = "Sanitation Access (%population)", ylab = "Sanitation(%)", xlab = "Year") %>%
  dyHighlight(highlightCircleSize = 5, 
              highlightSeriesBackgroundAlpha = 0.2,
              hideOnMouseOut = FALSE)

```

## Difference between Income Groups
Use the World Bank API to retrieve the safely managed sanitation facilities (% of population with access) data for all countries, starting from 1960. 

I segmented the dataset into 4 different income groups - LIC, LMC, UMC, and HIC.
General idea:

1. Use code 'SH.STA.SMSS.ZS' to get safely managed sanitation facilities (% of population with access) data. 
2. Use code 'SP.POP.TOTL' to get population data.
3. Drop the rows with missing population data or missing sanitation data.
5. Use weighted arithmetic mean to calculate the sanitation facilities (% population) for different income level.

$$sanitation(\%) = \frac{\sum\limits_{i=1}^{n} \left(population_{i} * sanitation(\%)_{i}\right)} {\sum\limits_{i=1}^{n} population_{i}}$$
   
For the 4 income groups, the trend of safely managed sanitation facilities (% of population with access):

1. Low-income economies (LIC), raised from 13.10% (2000) to 14.90% (2015), CAGR = 0.86%
2. Lower-middle-income economies (LMC), raised from 45.43% (2000) to 53.60% (2015), CAGR = 1.11%
3. Upper-middle-income economies (UMC), raised from 28.54% (2000) to 52.70% (2015), CAGR = 4.17%
4. High-income economies (HIC), raised from 83.78% (2000) to 87.84% (2015), CAGR = 0.31%

The Upper-middle-income economies has the highest compound annual growth rate (CAGR), 4.17%.
